---
title: "Vignette package rdadapt v1.0"
author: "Thibaut Capblancq & Maya Gueguen"
date: "2024"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette package rdadapt v1.0}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br>

`rdadapt` was developed to facilitate the exploration and projection of the gene \~ environment relationship across current and future landscapes. The package provides various statistical tools to conduct **landscape genomic analyses** going from *identifying the genetic bases of adaptation within a species* to *predict future population maladaptation to climate change*.

In the following, we show how the `rdadapt` package can perform:

+ variance partitioning,
+ genome scans for selection,
+ adaptive index projection on the landscape,
+ as well as estimation of genomic offsets
+ and identification of co-varying modules of adaptive loci or discrete adaptive groups of individual or populations.

*These analyses were inspired from various studies:*

- [Capblancq et al. 2018](http://dx.doi.org/10.1101/849406)
- [Forester et al. 2018]()
- [Capblancq et al. 2020](http://dx.doi.org/10.1146/annurev-ecolsys-020720-042553)
- [Capblancq & Forester 2021](http://dx.doi.org/10.1111/2041-210x.13722)
- [Capblancq et al. 2023](http://dx.doi.org/10.1111/nph.18465)
- [Hoste et al. 2023](http://dx.doi.org/10.1038/s41437-023-00661-2)


We show how to run the package using XXX.

To run the package, you need to install the package and load it using the following command lines:

```{r, message=FALSE, results='hide', eval=TRUE, warning=FALSE}
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("qvalue")
library(qvalue)

# library(devtools)
# install_github("landscape-genomics/rdadapt")
library(rdadapt)

library(vegan)
library(robust)
library(ggplot2)
library(terra)
library(foreach)
library(plyr)
library(viridis)
library(factoextra)
library(reshape2)
```

<br>

<h2>1. Data Preparation</h2>

<br>

`maladapt` dataset contains informations about individuals and populations, and corresponding genetic informations.

```{r, message=FALSE, warning=FALSE, eval=TRUE}
data("maladapt")
names(maladapt)
```

<br>

<h4>1.1 Metadata individuals, populations and genetic dataset</h4>

```{r, message=FALSE, warning=FALSE, eval=TRUE}
## Get population labels
pop <- maladapt$pop

## Get individual coordinates
coord <- maladapt$coord
plot(coord, cex = 0.2, xlim = c(0, 12), ylim = c(0, 12))

## Loading genetic dataset
Y <- maladapt$Y
Y[1:5, 1:5]
```

```{r, message=FALSE, warning=FALSE, eval=TRUE}
## Loading the current and future climatic variables
xx.current <- maladapt$xx.current
xx.future <- maladapt$xx.future

# Rasterize the variables
xx.ras.current <- rast(data.frame(x = rep(1:12, time = 12)
                                  , y = rep(1:12, each = 12)
                                  , aggregate(xx.current, by = list(ceiling(coord$x)
                                                                    , ceiling(coord$y))
                                              , mean)[, -c(1, 2)])
                       , type = "xyz"
                       , crs = crs("epsg:4326"))
plot(xx.ras.current)

xx.ras.future <- rast(data.frame(x = rep(1:12, time = 12)
                                 , y = rep(1:12, each = 12)
                                 , aggregate(xx.future, by = list(ceiling(coord$x)
                                                                  , ceiling(coord$y))
                                           , mean)[, -c(1, 2)])
                      , type = "xyz"
                      , crs = crs("epsg:4326"))
plot(xx.ras.future)
```

<h4>1.2 Subsampling populations</h4>

```{r, message=FALSE, eval=TRUE, fig.align='center'}
## Randomly select only 25 populations to mimic more what is in empirical datasets
samp <- which(pop %in% sample(1:100, 25, replace = FALSE))

## Subset the data
coord.samp <- coord[samp, ]
plot(coord.samp, cex = 0.2, xlim = c(0, 12), ylim = c(0, 12), col = "red")

Y.samp <- Y[samp, ]
xx.current.samp <- xx.current[samp, ]
xx.future.samp <- xx.future[samp, ]
```

<br>

<h2>2. Genome Scan</h2>

<br>

<h4>2.1 Identifying loci under selection using RDA</h4>

```{r, message=FALSE, eval=TRUE}
RDA_env <- rda(Y ~ .,  xx.current)
plot(RDA_env)
```

We then have to *choose a number of RDA axes to include* when conducting the genome scan.

```{r, message=FALSE, eval=TRUE, fig.align='center'}
screeplot(RDA_env, main = "Eigenvalues of constrained axes")
```

Looking at the proportion of variance explained by each axis, we decided to use the first two axes.

```{r, message=FALSE, results='hide', eval=TRUE}
## Running the function with K = 2
rdadapt_env <- rdadapt(RDA = RDA_env, K = 2)
```

One critical step when conducting a genome scan is to *set a pertinent p-value threshold to identify the outlier loci*. Here, we used a *Bonferroni correction* to account for multiple testing.

```{r, message=FALSE, eval=TRUE}
## P-values threshold after Bonferroni correction
thres_env <- 0.01 / length(rdadapt_env$p.values)

## Identifying the loci that are below the p-value threshold
outliers <- (1:ncol(Y))[which(rdadapt_env$p.values < thres_env)]
length(outliers)
```

The function found 215 outliers.

```{r, message=FALSE, eval=TRUE}
## Get adaptive loci for env variable 1 and env variable 2
causal1 <- maladapt$causal1
causal2 <- maladapt$causal2

## True positive ratio
sum(outliers %in% causal1 | outliers %in% causal2) / length(outliers)

## False negative ratio
sum(!(causal1 %in% outliers)) + sum(!(causal2 %in% outliers)) / (length(causal1) + length(causal2))

```

Once the outliers have been identified, it can be useful to visualize their distribution in comparison with neutral loci, using either a RDA biplot or a Manhattan plot.

```{r, message=FALSE, warning=FALSE, results='hide', eval=TRUE, fig.align='center'}
# ## Formatting table for ggplot
# locus_scores <- scores(RDA_env, choices=c(1:2), display="species", scaling="none") # vegan references "species", here these are the loci
# TAB_loci <- data.frame(names = row.names(locus_scores), locus_scores)
# TAB_loci$type <- "Neutral"
# TAB_loci$type[TAB_loci$names%in%outliers$Loci] <- "All outliers"
# TAB_loci$type[TAB_loci$names%in%outliers_rdadapt_env] <- "Top outliers"
# TAB_loci$type <- factor(TAB_loci$type, levels = c("Neutral", "All outliers", "Top outliers"))
# TAB_loci <- TAB_loci[order(TAB_loci$type),]
# TAB_var <- as.data.frame(scores(RDA_env, choices=c(1,2), display="bp")) # pull the biplot scores
# 
# ## Biplot of RDA loci and variables scores
# ggplot() +
#   geom_hline(yintercept=0, linetype="dashed", color = gray(.80), size=0.6) +
#   geom_vline(xintercept=0, linetype="dashed", color = gray(.80), size=0.6) +
#   geom_point(data = TAB_loci, aes(x=RDA1*20, y=RDA2*20, colour = type), size = 1.4) +
#   scale_color_manual(values = c("gray90", "#F9A242FF", "#6B4596FF")) +
#   geom_segment(data = TAB_var, aes(xend=RDA1, yend=RDA2, x=0, y=0), colour="black", size=0.15, linetype=1, arrow=arrow(length = unit(0.02, "npc"))) +
#   geom_text(data = TAB_var, aes(x=1.1*RDA1, y=1.1*RDA2, label = row.names(TAB_var)), size = 2.5, family = "Times") +
#   xlab("RDA 1") + ylab("RDA 2") +
#   facet_wrap(~"RDA space") +
#   guides(color=guide_legend(title="Locus type")) +
#   theme_bw(base_size = 11, base_family = "Times") +
#   theme(panel.background = element_blank(), legend.background = element_blank(), panel.grid = element_blank(), plot.background = element_blank(), legend.text=element_text(size=rel(.8)), strip.text = element_text(size=11))
# 
# ## Manhattan plot
# Outliers <- rep("Neutral", length(colnames(AllFreq)))
# Outliers[colnames(AllFreq)%in%outliers$Loci] <- "All outliers"
# Outliers[colnames(AllFreq)%in%outliers_rdadapt_env] <- "Top outliers"
# Outliers <- factor(Outliers, levels = c("Neutral", "All outliers", "Top outliers"))
# TAB_manhatan <- data.frame(pos = 1:length(colnames(AllFreq)), 
#                            pvalues = rdadapt_env$p.values, 
#                            Outliers = Outliers)
# TAB_manhatan <- TAB_manhatan[order(TAB_manhatan$Outliers),]
# ggplot(data = TAB_manhatan) +
#   geom_point(aes(x=pos, y=-log10(pvalues), col = Outliers), size=1.4) +
#   scale_color_manual(values = c("gray90", "#F9A242FF", "#6B4596FF")) +
#   xlab("Loci") + ylab("-log10(p.values)") +
#   geom_hline(yintercept=-log10(thres_env), linetype="dashed", color = gray(.80), size=0.6) +
#   facet_wrap(~"Manhattan plot", nrow = 3) +
#   guides(color=guide_legend(title="Locus type")) +
#   theme_bw(base_size = 11, base_family = "Times") +
#   theme(legend.position="right", legend.background = element_blank(), panel.grid = element_blank(), legend.box.background = element_blank(), plot.background = element_blank(), panel.background = element_blank(), legend.text=element_text(size=rel(.8)), strip.text = element_text(size=11))
```

<br>

<h2>3. Adaptive Index</h2>

<br>

<h4>3.1 Adaptively enriched genetic space</h4>

The 91 putative adaptive loci can be used as multivariate response in a new “adaptively enriched” RDA, using the 9 selected bioclimatic variables as explanatory variables.

```{r, message=FALSE, eval=TRUE}
## Adaptively enriched RDA
RDA_outliers <- rda(Y[, outliers] ~ .,  xx.current)
```

A RDA biplot allows us to visualize the relationship between the 91 putative adaptive loci and the underlying climatic variables.

```{r, message=FALSE, warning=FALSE, results='hide', eval=TRUE, fig.align='center', fig.height = 3, fig.width= 3}
TAB_loci <- scores(RDA_outliers, choices = c(1:2), display = "species", scaling="none")
TAB_var <- scores(RDA_outliers, choices = c(1:2), display = "bp")

## RDA biplot
ggplot() +
  geom_hline(yintercept = 0, linetype = "dashed", color = gray(.80), size = 0.6) +
  geom_vline(xintercept = 0, linetype = "dashed",  color = gray(.80), size = 0.6) +
  geom_point(data = TAB_loci
             , aes(x = RDA1 * 3, y = RDA2 * 3)
             , colour = "#EB8055FF", size = 2, alpha = 0.8) +
  geom_segment(data = TAB_var
               , aes(xend = RDA1, yend = RDA2, x = 0, y = 0)
               , colour = "black", size = 0.15, linetype = 1
               , arrow = arrow(length = unit(0.02, "npc"))) +
  geom_text(data = TAB_var
            , aes(x = 1.1 * RDA1, y = 1.1 * RDA2, label = row.names(TAB_var))
            , size = 2.5, family = "Times") +
  labs(x = "RDA 1 (XX%)", y = "RDA 2 (XX%)", subtitle = "Adaptively enriched RDA space") +
  theme_bw(base_size = 11, base_family = "Times") +
  theme(panel.grid = element_blank()
        , plot.background = element_blank()
        , panel.background = element_blank())
```

<br>

<h4>3.2 Adaptive index across the landscape</h4>

The scores of the environmental variables along the RDA axes can be used to calculate a genetic-based index of adaptation for each environmental pixel of the landscape. This index is estimated independently for each RDA axis of interest using the formula: $$
\sum_{i = 1}^{n}{a_ib_i}
$$ Where *a* is the climatic variable score (loading) along the RDA axis, *b* is the standardized value for this particular variable at the focal pixel, and *i* refers to one of the *n* different variables used in the RDA model.

```{r, message=FALSE, results='hide', eval=TRUE}
## Running the function for all pixels 
res_RDA_proj_current <- adaptive_index(RDA = RDA_outliers
                                       , K = 2
                                       , env = xx.ras.current
                                       , env_mask = NULL
                                       , method = "loadings")
```

The adaptive index thus provides an estimate of adaptive genetic similarity or difference of all pixels on the landscape as a function of the values of the environmental predictors at that location. When projected on a map it allows visualizing the different adaptive gradients across a species range.

```{r, message=FALSE, warning=FALSE, results='hide', eval=TRUE, fig.align='center', fig.width=10}
TAB_RDA <- as.data.frame(res_RDA_proj_current, xy = TRUE)
for (i in 3:ncol(TAB_RDA)) {
  TAB_RDA[, i] <- (TAB_RDA[, i] - min(TAB_RDA[, i])) / (max(TAB_RDA[, i]) - min(TAB_RDA[, i]))
}
TAB_RDA <- melt(TAB_RDA, c("x", "y"))
TAB_RDA$value_cut <- cut(TAB_RDA$value, breaks = seq(0, 1, length.out = 10), include.lowest = TRUE)


ggplot(data = TAB_RDA, aes(x = x, y = y, fill = value_cut)) + 
  geom_raster() +
  facet_grid(~ variable) +
  scale_fill_viridis_d("Adaptive index", alpha = 0.8, direction = -1, option = "A"
                       , labels = c("Negative scores", "", "", ""
                                    , "Intermediate scores", "", "", ""
                                    , "Positive scores")) +
  labs(x = "Longitude", y = "Latitude") +
  theme_bw(base_size = 11, base_family = "Times") +
  theme(panel.grid = element_blank()
        , plot.background = element_blank()
        , panel.background = element_blank()
        , strip.text = element_text(size = 11))
```

<br>

<h2>4. Genomic Offset</h2>

<br>

<h4>4.1 Predicting population future maladaptation</h4>

Once the genetic \~ environment relationship is characterized, it can be extrapolated to future environments to predict a potential shift in adaptive optimum induced by climate change.The RDA-based method to predict this future maladaptation is relatively simple. As done above, RDA can be used to predict the optimal adaptive genetic composition for each environmental pixel under consideration (Adaptive Index in Section 5), using both current and future environmental conditions. The difference between the two predictions provides an estimate of the change in genetic composition that would be required to track climate change.

```{r, message=FALSE, warning=FALSE, results='hide', eval=TRUE, fig.align='center', fig.width=10}
## Projecting the adaptive landscape after the change in environment 
res_RDA_proj <- genomic_offset(RDA = RDA_outliers
                               , K = 2
                               , env_pres = xx.ras.current
                               , env_fut = xx.ras.future
                               , env_mask = NULL
                               , method = "loadings")

## Table global genetic offset
RDA_proj_offset <- as.data.frame(res_RDA_proj$Proj_offset_global, xy = TRUE)

## Projecting genomic offset on a map
library(wesanderson)
colors <- wes_palette("Zissou1", 100, type = "continuous")

ggplot(data = RDA_proj_offset) + 
  geom_raster(aes(x = x, y = y, fill = Global_offset), alpha = 1) + 
  scale_fill_gradientn("Genomic offset", colours = colors) +
  labs(x = "Longitude", y = "Latitude") +
  theme_bw(base_size = 11, base_family = "Times") +
  theme(panel.grid = element_blank()
        , plot.background = element_blank()
        , panel.background = element_blank())
```

```{r, message=FALSE, warning=FALSE, results='hide', eval=TRUE, fig.align='center', fig.width=10}
## Verifying that prediction are matching with a decline in fitness on the landscape
diff.env <- sqrt(rowSums(as.data.frame((xx.ras.current[[1:2]] - xx.ras.future[[1:2]])^2)))
plot(diff.env, RDA_proj_offset$Global_offset)

## Get 
fitness <- maladapt$fitness
fitness.future <- rast(data.frame(x = rep(1:12, time = 12)
                                  , y = rep(1:12, each = 12)
                                  , aggregate(fitness[, 2], by = list(ceiling(coord$x)
                                                                     , ceiling(coord$y))
                                              , mean)[, -c(1, 2)])
                       , type = "xyz"
                       , crs = crs("epsg:4326"))
plot(log(fitness.future[]) ~ RDA_proj_offset$Global_offset)

```

<br>

<h4>4.2. Predicting geographic genomic offset</h4>

A challenge associated with any predictive method is to find ways to validate the predictions. Here, we estimated a geographic genomic offset between the climate of each source population and the climate of a hypothetical garden (cell 1,1).

```{r, message=FALSE, warning=FALSE, results='hide', eval=TRUE, fig.align='center', fig.height = 3.5, fig.width= 9}
## Estimates genomic offset for specific populations
offset_samp <- genomic_offset(RDA = RDA_outliers
                              , K = 2
                              , env_pres = xx.current.samp
                              , env_fut = xx.future.samp
                              , method = "loadings")

## Visualization
TAB_OFF <- data.frame(coord.samp, offset_samp$genomic_offset)
TAB_OFF <- melt(TAB_OFF, id.vars = c("x","y"))

ggplot() +
  geom_point(data = TAB_OFF, aes(x = x, y = y, color = value)) +
  facet_wrap( ~ variable) +
  scale_color_viridis("Genomic offset", direction = -1, alpha = 0.8) +
  theme_bw(base_size = 12, base_family = "Times") +
  theme(panel.grid = element_blank()
        , plot.background = element_blank()
        , panel.background = element_blank()
        , strip.text = element_text(size = 11))
```
```{r, message=FALSE, warning=FALSE, results='hide', eval=TRUE, fig.align='center', fig.height = 3.5, fig.width= 9}
## Estimates genomic offset for a transplantation experiment
offset_gar <- genomic_offset(RDA = RDA_outliers
                             , K = 2
                             , env_pres = xx.current
                             , env_gar = xx.current[100, ]
                             , method = "loadings")

## Visualization
TAB_OFF <- data.frame(coord, offset_gar$genomic_offset)
TAB_OFF <- melt(TAB_OFF, id.vars = c("x","y"))

ggplot() +
  geom_point(data = TAB_OFF, aes(x = x, y = y, color = value)) +
  geom_point(data = coord[100, ],  aes(x = x, y = y)
             , colour = "black", size = 3, shape = 4) +
  facet_wrap( ~ variable) +
  scale_color_viridis("Genomic offset", direction = -1, alpha = 0.8, option = "magma") +
  theme_bw(base_size = 12, base_family = "Times") +
  theme(panel.grid = element_blank()
        , plot.background = element_blank()
        , panel.background = element_blank()
        , strip.text = element_text(size = 11))
```

Based on the adaptively enriched RDA space, we estimated a genomic offset between each source individual and the common garden climatic conditions.


This geographic genomic offset can then be compared to different fitness traits in the garden to estimate the impact of this predicted maladaptation on seedling fitness. We can also compare the influence of genomic offset on fitness with the impact of a basic climate transfer distance estimated between each source population and the garden from the climatic variables only (using Mahalanobis distance).

<br>

<h2>5. Loci Modules</h2>

<br>

We can find covarying sets of loci on the RDA space

```{r, message=FALSE, results='hide', eval=TRUE, fig.align='center', fig.align='center', fig.height = 3.5, fig.width= 5}
## Groups the outlier loci into two modules
modules <- loci_modules(RDA = RDA_outliers, nb_clusters = 4)

## Visualization
ggplot() +
  geom_polygon(data = modules$polygons
               , aes(x = x, y = y, colour = factor(cluster)
                     , fill = factor(cluster)), alpha = 0.2) +
  geom_point(data = modules$loci, aes(x = x, y = y, colour = factor(cluster)
                                      , fill = factor(cluster)), size = 2, alpha = 0.7) +
  scale_fill_viridis(guide = "none", discrete = TRUE, direction = -1, alpha = 0.2) +
  scale_color_viridis(guide = "none", discrete = TRUE, direction = -1, alpha = 0.7) +
  labs(x = "Absolute score on RDA1", y = "Absolute score on RDA2", subtitle = "Modules of outlier loci") +
  theme_bw(base_size = 12, base_family = "Times") +
  theme(panel.grid = element_blank()
        , plot.background = element_blank()
        , panel.background = element_blank()
        , strip.text = element_text(size = 11))
```

<br>

<h2>6. Adaptive groups</h2>

<br>

The goal here is to identify groups of individuals with the same adaptive background


```{r, message=FALSE, results='hide', eval=TRUE, fig.align='center'}
## New RDA model with only sample populations and outlier loci
RDA_samp <- rda(Y.samp[,outliers] ~ .,  xx.current.samp)

## Finds N optimal groups among the individuals used to build the adaptive RDA space
groups <- adaptive_groups(RDA = RDA_samp, K = 2, nb_clusters = 2)

## Visualization in the RDA space
ggplot() +
  geom_polygon(data = groups$polygons
               , aes(x = x, y = y, colour = factor(cluster)
                     , fill = factor(cluster)), alpha = 0.2) +
  geom_point(data = groups$samples
             , aes(x = x, y = y, colour = factor(cluster)
                   , fill = factor(cluster)), size = 2, alpha = 0.7) +
  scale_fill_viridis(guide = "none", discrete = TRUE, direction = -1, alpha = 0.2) +
  scale_color_viridis(guide = "none", discrete = TRUE, direction = -1, alpha = 0.7) +
  labs(x = "Score on RDA1", y = "Score on RDA2", subtitle = "Adaptive groups") +
  theme_bw(base_size = 12, base_family = "Times") +
  theme(panel.grid = element_blank()
        , plot.background = element_blank()
        , panel.background = element_blank()
        , strip.text = element_text(size = 11))

## Visualization on the landscape
tab <- data.frame(coord.samp, cluster = groups$samples$cluster)
plot(tab[, 1:2], col = tab$cluster)

```

```{r, message=FALSE, results='hide', eval=TRUE, fig.align='center'}
## Finds N groups on the adaptive landscape projected from the RDA model
groups <- adaptive_groups(RDA = RDA_samp, K = 2, env = xx.ras.current, nb_clusters = 9)

## Visualization in the RDA space
ggplot() +
  geom_polygon(data = groups$polygons
               , aes(x = x, y = y, colour = factor(cluster)
                     , fill = factor(cluster)), alpha = 0.2) +
  geom_point(data = groups$samples
             , aes(x = x, y = y, colour = factor(cluster)
                   , fill = factor(cluster)), size = 2, alpha = 0.7) +
  scale_fill_viridis(guide = "none", discrete = TRUE, direction = -1, alpha = 0.2) +
  scale_color_viridis(guide = "none", discrete = TRUE, direction = -1, alpha = 0.7) +
  labs(x = "Score on RDA1", y = "Score on RDA2", subtitle = "Adaptive groups") +
  theme_bw(base_size = 12, base_family = "Times") +
  theme(panel.grid = element_blank()
        , plot.background = element_blank()
        , panel.background = element_blank()
        , strip.text = element_text(size = 11))
```
